Handling Recursion in Generic Programming Using Closed Type Families

### 2018-05-10 lncs-paper.tex
Replace  with

For chaining moves and edits, we can follow Yakushev~et~al.~\cite{MuRec2009} and employ  the flipped function composition \comp~and Kleisli composition \kleisli. The latter one is instantiated with the \texttt{Maybe} monad that wraps the result type of the movement functions.
\texttt{
\vs
\indent (\comp) \ann (a \ra b) \ra (b \ra c) \ra (a \ra c)\\
\indent (\kleisli) \ann Monad m \Ra (a \ra m b) \ra (b \ra m c) \ra (a \ra m c)
\vspace{0.2cm}
}

@@

The location structure consists of a focal subtree, which is one of the mutually recursive nodes of the whole structure of the family of datatypes, and its surrounding context:
\texttt{
\vs
\indent data Loc (r \ann *) (fam \ann [*]) (c \ann * \ra Constraint) where\\
\indent\s\s Loc \ann\hs Family r a fam c \ra Contexts r a fam c\\
\indent\ind\s\s\ra Loc r fam c
\vs
}
The type parameters \texttt{r}, \texttt{fam}, and \texttt{c} in \texttt{Loc} correspond to the root type of the tree, the list of types of nodes to visit, and a constraint imposing restrictions on the types in the list, respectively. Also, the single constructor is existentially quantified over one more type variable, \texttt{a}, for we need to store a type of the focus' parent to be able to move up successively in a tree-like structure. We discuss both term parameters of the constructor of \texttt{Loc} in detail below.

\subsubsection{Families}

The subtree in a focus is wrapped by the \texttt{Family} datatype. 
The wrapper incapsulates the proofs about a number of important properties of a focus.
\texttt{
\vs
\indent data Family (r \ann *) (a \ann *) (fam \ann [*])\\
\indent\ind\ind\ind (c \ann * \ra Constraint) where\\
\indent\s\s Family \ann\hs (Generic b, In b fam, ZipperI r a b fam c)\\
\indent\ind\ind\s\Ra b \ra Family r a fam c
\vs
}
Existential type variable \texttt{b} represents the type of a focus. We apply a number of predicates to \texttt{b}, hence we can implement the steps of the navigation not knowing the actual type of a focus. Firstly, the type of a focus should have the \texttt{Generic} representation. Secondly, it should live \texttt{In} the list of types we are going to visit. Lastly, it ought to satisfy the set of constraints for the whole zipper interface captured by the \texttt{ZipperI} predicate. In particular, the predicate ensures that \texttt{a} is the type of the parent for the focus in the structure under consideration.

@@

One last missing piece for managing focuses is the class \texttt{ProofIn}. It provides a proof of membership of a focus type to a family. Again, this generalizes the proof of type equality from Section~\ref{sec:handling-recursion}. The definitions of \texttt{ProofIn} and an auxiliary class \texttt{ProofFam} are given in Figure~\ref{fig:proof-in}. 
%The constraints on the second \texttt{ProofFam} instance, which repeat the same ones from the \texttt{Family} constructor, constitute the proof.
% this is not clear and, I think, can be omitted

\begin{figure}[t]
\centering
\normalsize
\begin{tabular}{l}
\tt class ProofFam (inFam \ann Bool) (r \ann *) (a \ann *) (b \ann *)\\
\tt\ind\ind\ind\s\s\s (fam \ann [*]) (c \ann * \ra Constraint) where\\
\tt\s\s witness \ann b \ra Maybe (Family r a fam c)
\vs
\tt instance ProofFam 'False r a b fam c where\\
\tt\s\s witness \_ \hs\nths= Nothing\\
\tt instance (Generic b, In b fam, ZipperI r a b fam c)\\
\tt\ind\Ra ProofFam 'True r a b fam c where\\
\tt\s\s witness\s\s\s = Just . Family
\vs
\tt class ProofFam (InFam b fam) r a b fam c\\
\tt\ind\Ra ProofIn r a b fam c\\
\tt instance ProofFam (InFam b fam) r a b fam c\\
\tt\ind\Ra ProofIn r a b fam c
\end{tabular}
\caption{Proof of membership of a family of datatypes.}
\label{fig:proof-in}
\end{figure}

@@

When we focus on a particular node in a data structure, having the surrounding context of that node is enough to reconstruct the entire structure. The context for the tree's location has the shape of the original structure of that tree but with one hole at the place of its focus, whence it is sometimes called a \emph{one-hole context}. The context can be expressed as a stack (\texttt{Contexts} below), where each frame \texttt{Context} corresponds to the particular node with a hole: it ascends from the focal node keeping its siblings, the siblings of its parent, and until it reaches the root node, through all levels in the tree. So the stack of contexts essentially reflects the track of the movement in the structure.
\texttt{
\vs
\indent data Contexts (r \ann *) (a \ann *) (fam \ann [*])\\
\indent\ind\ind\ind\s\s (c \ann * \ra Constraint) where\\
\indent\s\s CNil \ann\hs Contexts a a fam c\\
\indent\s\s Ctxs \ann\hs (Generic a, In a fam, ZipperI r x a fam c)\\
\indent\ind\s\s\s\Ra Context fam a \ra Contexts r x fam c\\
\indent\ind\s\s\s\ra Contexts r a fam c
\vs
}
This has the shape similar to the previously defined \texttt{Loc} and \texttt{Family} datatypes: it keeps the types of the root and of the immediate parent of the focal node (of the union with the focal subtree when that plugs the hole)---as ensured by sharing that type with \texttt{Family} in the \texttt{Loc} constructor. The \texttt{ZipperI} constraint with the type \texttt{x} of the previous context frame here indicates that the constraint for the zipper holds after plugging the hole, so all the properties, conversely, can be proved by induction for the focus type when it moves down in the tree adding new contexts onto the stack---and the \texttt{CNil} constructor for an empty context with the \texttt{r} and \texttt{a} types equal is the inductive basis in that proof. Note that the type of the current focus is not reflected in the \texttt{Contexts} datatype, as we do not need that.

@@

To move focus down to the leftmost child of the current focal node in the tree, we should analyze the focal subtree's representation to find its first immediate child, and compute its respective context. The following definition of \texttt{goDown} uses two auxiliary functions: \texttt{toFirst} and \texttt{toFirstCtx}.
\texttt{
\vs
\indent goDown \ann Loc a fam c \ra Maybe (Loc a fam c)\\
\indent goDown (Loc (Family t) cs)\\
\indent\s\s = case toFirst t of\\
\indent\ind\s\s Just t' \ra Just \$ Loc t' (Ctxs (toFirstCtx t) cs)\\
\indent\ind\s\s\_\ind\s\s\s\hs\ra Nothing
\vs
}
The function \texttt{toFirst} returns its result in the \texttt{Maybe} monad, and may return \texttt{Nothing}, if the focal node has no children, that is, we currently focus on the leaf node and cannot go down. The function \texttt{toFirstCtx} should not fail: if we can move, it computes the context that matches the leftmost subtree selected from the focus' children.

\begin{figure}[t]
\centering
\normalsize
\begin{tabular}{l}
\tt toFirst \ann\hs\fa fam c r a. (Generic a, ToFirst r a fam c)\\
\tt\ind\ind\Ra a \ra Maybe (Family r a fam c)\\
\tt toFirst t = appToNP @AllProof toFirstNP \$ unSOP \$ from t
\vspace{0.3cm}\\
\bf Proof
\vs
\tt class\ind All (ProofIn r a fam c) xs \Ra AllProof r a fam c xs\\
\tt instance All (ProofIn r a fam c) xs \Ra AllProof r a fam c xs
\vs
\tt type ToFirst r a fam c = All (AllProof r a fam c) (Code a)
\vspace{0.3cm}\\
\bf Processing products
\vs
\tt toFirstNP \ann\hs\fa fam c r a xs. All (ProofIn r a fam c) xs\\
\tt\ind\ind\s\s\Ra NP I xs \ra Maybe (Family r a fam c)\\
\tt toFirstNP (I (x \ann b) :* xs)\\
\tt\s\s = witness @(InFam b fam) x `mplus` toFirstNP xs\\
\tt toFirstNP Nil = Nothing
\end{tabular}
\caption{Implementation of \texttt{toFirst}.}
\label{fig:toFirst}
\end{figure}

>>>>

For chaining moves and edits, we can follow Yakushev~et~al.~\cite{MuRec2009} and employ  the flipped function composition \comp~and Kleisli composition \kleisli. The latter is instantiated with the \texttt{Maybe} monad that wraps the result type of the movement functions.
\texttt{
\vs
\indent (\comp) \ann (a \ra b) \ra (b \ra c) \ra (a \ra c)\\
\indent (\kleisli) \ann Monad m \Ra (a \ra m b) \ra (b \ra m c) \ra (a \ra m c)
\vspace{0.2cm}
}

@@

The location structure consists of a focal subtree, which is one of the mutually recursive nodes of the whole structure of the family of datatypes, and its surrounding context:
\texttt{
\vs
\indent data Loc (r \ann *) (fam \ann [*]) (c \ann * \ra Constraint) where\\
\indent\s\s Loc \ann\hs Focus r a fam c \ra Contexts r a fam c\\
\indent\ind\s\s\ra Loc r fam c
\vs
}
The type parameters \texttt{r}, \texttt{fam}, and \texttt{c} in \texttt{Loc} correspond to the root type of the tree, the list of types of nodes to visit, and a constraint imposing restrictions on the types in the list, respectively. Also, the single constructor is existentially quantified over one more type variable, \texttt{a}, for we need to store a type of the focus' parent to be able to move up successively in a tree-like structure. We discuss both term parameters of the constructor of \texttt{Loc} in detail below.

\subsubsection{Focus}

The subtree in a focus is wrapped by the \texttt{Focus} datatype. 
The wrapper incapsulates the proofs about a number of important properties of a focus.
\texttt{
\vs
\indent data Focus (r \ann *) (a \ann *) (fam \ann [*])\\
\indent\ind\ind\ind (c \ann * \ra Constraint) where\\
\indent\s\s Focus \ann\hs (Generic b, In b fam, ZipperI r a b fam c)\\
\indent\ind\ind\s\Ra b \ra Focus r a fam c
\vs
}
Existential type variable \texttt{b} represents the type of a focus. We apply a number of predicates to \texttt{b}, hence we can implement the steps of the navigation not knowing the actual type of a focus. Firstly, the type of a focus should have the \texttt{Generic} representation. Secondly, it should live \texttt{In} the list of types we are going to visit. Lastly, it ought to satisfy the set of constraints for the whole zipper interface captured by the \texttt{ZipperI} predicate. In particular, the predicate ensures that \texttt{a} is the type of the parent for the focus in the structure under consideration.

@@

One last missing piece for managing focuses is the class \texttt{ProofIn}. It provides a proof of membership of a focus type to a family. Again, this generalizes the proof of type equality from Section~\ref{sec:handling-recursion}. The definitions of \texttt{ProofIn} and an auxiliary class \texttt{ProofFocus} are given in Figure~\ref{fig:proof-in}. 
%The constraints on the second \texttt{ProofFam} instance, which repeat the same ones from the \texttt{Family} constructor, constitute the proof.
% this is not clear and, I think, can be omitted

\begin{figure}[t]
\centering
\normalsize
\begin{tabular}{l}
\tt class ProofFocus (inFam \ann Bool) (r \ann *) (a \ann *) (b \ann *)\\
\tt\ind\ind\ind\s\s\s (fam \ann [*]) (c \ann * \ra Constraint) where\\
\tt\s\s witness \ann b \ra Maybe (Focus r a fam c)
\vs
\tt instance ProofFocus 'False r a b fam c where\\
\tt\s\s witness \_ \hs\nths= Nothing\\
\tt instance (Generic b, In b fam, ZipperI r a b fam c)\\
\tt\ind\Ra ProofFocus 'True r a b fam c where\\
\tt\s\s witness\s\s\s = Just . Focus
\vs
\tt class ProofFocus (InFam b fam) r a b fam c\\
\tt\ind\Ra ProofIn r a b fam c\\
\tt instance ProofFocus (InFam b fam) r a b fam c\\
\tt\ind\Ra ProofIn r a b fam c
\end{tabular}
\caption{Proof of membership of a family of datatypes.}
\label{fig:proof-in}
\end{figure}

@@

When we focus on a particular node in a data structure, having the surrounding context of that node is enough to reconstruct the entire structure. The context for the tree's location has the shape of the original structure of that tree but with one hole at the place of its focus, whence it is sometimes called a \emph{one-hole context}. The context can be expressed as a stack (\texttt{Contexts} below), where each frame \texttt{Context} corresponds to the particular node with a hole: it ascends from the focal node keeping its siblings, the siblings of its parent, and until it reaches the root node, through all levels in the tree. So the stack of contexts essentially reflects the track of the movement in the structure.
\texttt{
\vs
\indent data Contexts (r \ann *) (a \ann *) (fam \ann [*])\\
\indent\ind\ind\ind\s\s (c \ann * \ra Constraint) where\\
\indent\s\s CNil \ann\hs Contexts a a fam c\\
\indent\s\s Ctxs \ann\hs (Generic a, In a fam, ZipperI r x a fam c)\\
\indent\ind\s\s\s\Ra Context fam a \ra Contexts r x fam c\\
\indent\ind\s\s\s\ra Contexts r a fam c
\vs
}
This has the shape similar to the previously defined \texttt{Loc} and \texttt{Focus} datatypes: it keeps the types of the root and of the immediate parent of the focal node (of the union with the focal subtree when that plugs the hole)---as ensured by sharing that type with \texttt{Focus} in the \texttt{Loc} constructor. The \texttt{ZipperI} constraint with the type \texttt{x} of the previous context frame here indicates that the constraint for the zipper holds after plugging the hole, so all the properties, conversely, can be proved by induction for the focus type when it moves down in the tree adding new contexts onto the stack---and the \texttt{CNil} constructor for an empty context with the \texttt{r} and \texttt{a} types equal is the inductive basis in that proof. Note that the type of the current focus is not reflected in the \texttt{Contexts} datatype, as we do not need that.

@@

To move focus down to the leftmost child of the current focal node in the tree, we should analyze the focal subtree's representation to find its first immediate child, and compute its respective context. The following definition of \texttt{goDown} uses two auxiliary functions: \texttt{toFirst} and \texttt{toFirstCtx}.
\texttt{
\vs
\indent goDown \ann Loc a fam c \ra Maybe (Loc a fam c)\\
\indent goDown (Loc (Focus t) cs)\\
\indent\s\s = case toFirst t of\\
\indent\ind\s\s Just t' \ra Just \$ Loc t' (Ctxs (toFirstCtx t) cs)\\
\indent\ind\s\s\_\ind\s\s\s\hs\ra Nothing
\vs
}
The function \texttt{toFirst} returns its result in the \texttt{Maybe} monad, and may return \texttt{Nothing}, if the focal node has no children, that is, we currently focus on the leaf node and cannot go down. The function \texttt{toFirstCtx} should not fail: if we can move, it computes the context that matches the leftmost subtree selected from the focus' children.

\begin{figure}[t]
\centering
\normalsize
\begin{tabular}{l}
\tt toFirst \ann\hs\fa fam c r a. (Generic a, ToFirst r a fam c)\\
\tt\ind\ind\Ra a \ra Maybe (Focus r a fam c)\\
\tt toFirst t = appToNP @AllProof toFirstNP \$ unSOP \$ from t
\vspace{0.3cm}\\
\bf Proof
\vs
\tt class\ind All (ProofIn r a fam c) xs \Ra AllProof r a fam c xs\\
\tt instance All (ProofIn r a fam c) xs \Ra AllProof r a fam c xs
\vs
\tt type ToFirst r a fam c = All (AllProof r a fam c) (Code a)
\vspace{0.3cm}\\
\bf Processing products
\vs
\tt toFirstNP \ann\hs\fa fam c r a xs. All (ProofIn r a fam c) xs\\
\tt\ind\ind\s\s\Ra NP I xs \ra Maybe (Focus r a fam c)\\
\tt toFirstNP (I (x \ann b) :* xs)\\
\tt\s\s = witness @(InFam b fam) x `mplus` toFirstNP xs\\
\tt toFirstNP Nil = Nothing
\end{tabular}
\caption{Implementation of \texttt{toFirst}.}
\label{fig:toFirst}
\end{figure}
